---
title: "EWAS in MrOS - Horvath analysis"
author: "Dan Evans"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---


```{r, setup}
knitr::opts_chunk$set(cache.lazy = FALSE)
```
# Load libraries
```{r}
library(xtable)
library(tidyverse)
library(multtest)
```

# Create input files for Horvath website

It turned out that the input files were fine, except for the line endings! I created the files in linux which produces files with line endings like \n, but the Horvath site expected windows style line endings (\r\n). `unix2dos` didn't solve the problem. The only way I could solve this was to open the file in Excel in terminal server, and save as csv file. This Excel trick was suggested on the horvath website. So hacky, but whatever. 


```{r, eval = FALSE}
library(data.table)
pheno <- fread("../data/pheno/INFLAME.CSV")
pheno <- pheno[,.(ID, V3AGE1)]
setnames(pheno, "V3AGE1", "Age")
#sample annotation in same order as samples in betaMat
samp <- fread("../data/raw/Evans_Project_002/Evans_Project_002_Sample_Sheet.csv", skip = 8)
samp[, index := seq_along(Sample_Name)]
sum(colnames(betaMat)!= samp$Sample_Name)
sum(colnames(betaMat)== samp$Sample_Name)
samp <- merge(samp, pheno, by.x = "Sample_Name", by.y = "ID", all.x = TRUE)
samp[, Tissue := "Blood WB"]
samp[, Female := 0L]
samp <- samp[order(index)]
setnames(samp, "Sample_Name", "ID")
samp <- samp[,.(ID, Age, Female, Tissue)]
samp[is.na(Age), Age := as.integer(mean(samp$Age, na.rm = TRUE))]
fwrite(samp, file = "../data/horvath/input/sampleAnnotation.csv", eol = "\r\n", na= "NA")

horvathProbe <- read.csv("../data/horvath/input/datMiniAnnotation3.csv", header=T, stringsAsFactors = F)#28587 probes
sum(!horvathProbe$Name %in% rownames(betaMat)) #Darn, 2415 probes not in my data

betasDF <- as.data.frame(betaMat)
betasDF$ProbeID <- row.names(betaMat)
betasDF <- betasDF[,c(ncol(betasDF), 1:ncol(betasDF)-1)]
match1 <- match(horvathProbe[,1], betasDF[,1])
betasDFreduced <- betasDF[match1,]
betasDFreduced[is.na(match1), 1] <- as.character(horvathProbe[is.na(match1), 1])

temp <- sapply(betasDFreduced, function(x) is.numeric(x))
sum(temp)
temp <- sapply(betasDFreduced, function(x) !is.numeric(x))
sum(temp)

write.csv(betasDFreduced, "../data/horvath/input/MethylData.csv", row.names = F, quote = F, eol = "\r\n")

```

# Horvath output analysis

```{r}

pheno <- read_csv("../data/pheno/INFLAME.CSV")
horvath <- read_csv("../data/horvath/output/noImputation/MethylData_noImputation.output.csv")

dat <- inner_join(horvath, pheno, by = c("SampleID" = "ID"))

p <- ggplot(data = dat, mapping = aes(x = Age, y= DNAmAge)) +
	geom_point() + 
	geom_smooth(method = "lm") +
	geom_abline()

print(p)

cor.test(dat$Age, dat$DNAmAge)
```

```{r, results = "asis"}
lm1 <- lm(log(CYCRPJH) ~ DNAmAge + Age + CD8.naive + CD8pCD28nCD45RAn + PlasmaBlast + CD4T + NK + Mono + Gran, data = dat)
print(xtable(lm1), type = "html")

lm1 <- lm(CYCRPJH ~ DNAmAge + Age + CD8.naive + CD8pCD28nCD45RAn + PlasmaBlast + CD4T + NK + Mono + Gran, data = dat)
print(xtable(lm1), type = "html")
```


# Probe-wise association analysis with inflammation
```{r, results = "asis", eval = FALSE}

phenoMrOS <- read_csv("../data/pheno/INFLAME.CSV")
horvath <- read_csv("../data/horvath/output/noImputation/MethylData_noImputation.output.csv")

pheno <- inner_join(phenoMrOS, horvath, by = c("ID" = "SampleID"))

#load betas data
betas <- read_csv("../data/clean/betas.csv")
samp <- read_csv("../data/raw/Evans_Project_002/Evans_Project_002_Sample_Sheet.csv", skip = 8)
samp <- samp %>%
	mutate(sample_ID = paste(Sentrix_ID, Sentrix_Position, sep = "_"))
#sample IDs in sample sheet in the same order as sample names in betas file
sum(names(betas)[-1] != samp$sample_ID)
sum(names(betas)[-1] == samp$sample_ID)
names(betas) <- c("probeID", samp$Sample_Name)
#create matrix of betas for apply
betaMat <- betas %>%
	select(-probeID) %>%
	as.matrix

betaMat <- betaMat[,colnames(betaMat)  %in% pheno$ID]
#remove trailing info from probe IDs
probes <- base::strsplit(betas$probeID, split = "_", fixed = TRUE)
probes <- vapply(probes, function(x) x[1], FUN.VALUE = character(1))
dimnames(betaMat)[[1]] <- probes

lm_fun <- function(cg){
  out <- tryCatch({
    lm1 <- lm(log(CYCRPJH) ~ cg + Age + CD8.naive + CD8pCD28nCD45RAn + PlasmaBlast + CD4T + NK + Mono + Gran, data = pheno)
    n1 <- length(resid(lm1))
    b <- summary(lm1)$coefficients[2,1]
    se <- summary(lm1)$coefficients[2,2]
    p <- summary(lm1)$coefficients[2,4]
    return(c(n1, b, se, p))
  },
  error = function(e) rep(NA, 4)
  )
}

results <- apply(betaMat, 1, lm_fun)
results <- t(results)
colnames(results) <- c("N", "Beta", "SE", "P")
resultsTB <- as_tibble(results)
resultsTB <- resultsTB %>%
	mutate(probeID = rownames(results)) %>%
	select(probeID, everything())
adjp1 <- mt.rawp2adjp(resultsTB[["P"]], proc = "BH")
resultsTB <- resultsTB %>%
	mutate(P_BH = adjp1$adjp[order(adjp1$index),"BH"])

write_csv(resultsTB, "../results/probewise_CRPlog.csv")
resultsDF_top <- resultsTB %>%
	arrange(P) %>%
	head(50) %>%
	as.data.frame

print(xtable(resultsDF_top), type = "html")


resultsTB_top <- resultsTB %>%
	arrange(P) %>%
	head(50)

print(xtable(resultsTB_top), type = "html")
```

