---
title: "EWAS pilot in MrOS using minfi R package"
author: "Dan Evans"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

# EWAS terms

More drugs to examine. Anti-NGF and might be associated with OA. Anti-sclerostin and heart disease. 

DNA methylation quantity is expressed as $\beta$ 

$$ \beta = M/(M + U)  $$

Where M = hybridization signal from a methylated version of a cytosine nucleotide and 
U =  hybridization signal from an unmethylated version of a cytosine nucleotide. 

A detection probability represents the probability of a detected signal being background flourescence. If the probability is high, the signal is more likely to be background, and the value should be set to missing.

Standard workflows suggest to remove data points with detection P-value > 0.05.  

# Processing Illumina EPIC array with Sesame package

- EWAS array Illumina EPIC 850 + custom content

- LifeEGX worked with Sesame developer to process chips. LifeEGX recommends Sesame

- Sesame is a [bioconductor package](https://www.bioconductor.org/packages/release/bioc/html/sesame.html)
  + Improvements on previous EWAS packages for low-level processing. 
  + Existing methods do not identify artifacts associated with detection failure. Sources of failure include: insufficient DNA due to germline or somatic deletions or hyperpolymorphism, probe cross-hybridization. 
  + P-value with out-of-band array hybridization: pOOBAH
  + Reduces technical artifacts

- Lots of sesame issues. First, estimateCellComposition function was not found. Strangely, I could access the help file for the function in R. Something wrong with loading the function. So, I tried updating sesame, but tons of errors updating the associated sesameData package. I finally got them both installed, after just trying a bunch of times, but estimateCellComposition function still not found. 

- I will try minfi. 

```{r, setup}
knitr::opts_chunk$set(cache.lazy = FALSE)
```

# Minfi installation on UCR cluster


```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("minfi", lib = "~/Rlibs")
```

# Load libraries
```{r}
library(minfi)
```

# Filter and normalize data

This process takes a long time, so run it once and save the output.

```{r, eval = FALSE}

manifest <- read.csv("../data/raw/Evans_Project_002/CombinedManifestEPICplus.manifest.LifeEGX.csv", header = T, stringsAsFactors = F, na.strings = c("NA", ""))

#manifest <- readRDS("../data/raw/Evans_Project_002/OpenDMAP_sesame_manifest.rds")



baseDir <- c("../data/raw/Evans_Project_002/idat_Files/")
list.files(baseDir)
list.files(file.path(baseDir, "203259070019"))

targets <- read.metharray.sheet(base = "../data/raw/Evans_Project_002/", pattern = "*Sample_Sheet.csv")

RGset <- read.metharray.exp(targets = targets)
pd <- pData(RGset)

dat <- preprocessNoob(RGset)

```

# Load matrix of betas

Add MrOS subject IDs.

Fix probe IDs to allow matching with various downstream utilities.

```{r, cache = TRUE}
betas <- fread("../data/clean/betas.csv")
samp <- fread("../data/raw/Evans_Project_002/Evans_Project_002_Sample_Sheet.csv", skip = 8)
samp[, sample_ID := paste(Sentrix_ID, Sentrix_Position, sep = "_")]
#sample IDs in sample sheet in the same order as sample names in betas file
sum(names(betas)[-1] != samp$sample_ID)
sum(names(betas)[-1] == samp$sample_ID)
setnames(betas, c("probeID", samp$Sample_Name))
betaMat <- as.matrix(betas[,!"probeID"])

#remove trailing info from probe IDs
probes <- base::strsplit(betas$probeID, split = "_", fixed = TRUE)
probes <- vapply(probes, function(x) x[1], FUN.VALUE = character(1))
dimnames(betaMat)[[1]] <- probes
betaMat <- t(betaMat)

```

# Use beta matrix to predict cell composition

```{r}
g <- getRefSet(platform = "EPIC")
probeCell <- row.names(g)
q <- betaMat[,1]
probes2 <- base::intersect(names(q), probeCell)
q <- q[names(q) %in% probes2]
g <- g[probeCell %in% probes2,]

cells <- estimateCellComposition(g, q, refine = FALSE)


?estimateCellComposition

```

# Probe hg19 annotation

```{r}
EPIC.hg19.manifest <- sesameDataGet("EPIC.hg19.manifest")
str(EPIC.hg19.manifest)

```

# Use beta matrix to predict biological age

```{r}

(ageH353 <- apply(betaMat, 1, predictAgeHorvath353))

```

# Missing analysis 

By sample and probe
```{r}
miss_sample <- apply(betaMat, 1, function(x) sum(is.na(x))/length(x) )
miss_probe <- apply(betaMat, 2, function(x) sum(is.na(x))/length(x) )
length(miss_sample)
length(miss_probe)
sum(is.na(miss_sample))
sum(is.na(miss_probe))
sum(miss_sample >= 0.95) #0 samples have missing rate greater than 
sum(miss_sample >= 0.05) #14 samples with missing rate greater than 5%
sum(miss_probe >= 1) #3229 completely missing probes. Remove these, then determine number of samples with missing rate>0.05.
blank_probes <- names(miss_probe[miss_probe >= 1])
betaMat <- betaMat[,!colnames(betaMat) %in% blank_probes]
sum(miss_probe >= 0.95) #6608 probes with missing rate greater than 95%
sum(miss_probe >= 0.05) #89822 probes with missing rate greater than 5%
sum(miss_probe >= 0.05) / length(miss_probe) #10.4% probes have greater than 5% missing  

miss_probe[miss_probe >= 0.95]
```

# Read IDATs into SigSet list
```{r, cache = TRUE, results = "asis"}

IDATprefixes <- searchIDATprefixes(dir.name = "../data/raw/Evans_Project_002/idat_Files")
#IDATprefixes <- IDATprefixes[2:3]
ssets <- lapply(IDATprefixes, readIDATpair)
chipSex <- sapply(ssets, inferSex)
chipEth <- sapply(ssets, inferEthnicity)

qc10 <- do.call(rbind, lapply(ssets, function(x)
    as.data.frame(sesameQC(x))))
qc10$sample_name <- names(ssets)

qctab <- qc10[,c('mean_beta_cg','frac_meth_cg','frac_unmeth_cg','sex','age')]
tab <- xtable(qctab)
print(tab, type = "html")
```


# Mean intensity

The mean {M,U} intensity can be reached by mean_intensity. Similarly, the mean M+U intensity can be reached by mean_intensity_total. Low intensities are symptomatic of low input or poor hybridization.

```{r, results = "asis"}

library(wheatmap)
p1 <- ggplot(qc10) +
    geom_bar(aes(sample_name, mean_intensity), stat='identity') +
    xlab('Sample Name') + ylab('Mean Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc10) +
    geom_bar(aes(sample_name, mean_intensity_total), stat='identity') +
    xlab('Sample Name') + ylab('Mean M+U Intensity') +
    ylim(0,18000) +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())

```

# Fraction of NA

The fraction of NAs are signs of masking due to variety of reasons including failed detection, high background, putative low quality probes etc. 

```{r, results = "asis"}
p1 <- ggplot(qc10) +
    geom_bar(aes(sample_name, num_na_cg), stat='identity') +
    xlab('Sample Name') + ylab('Number of NAs') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
p2 <- ggplot(qc10) +
    geom_bar(aes(sample_name, frac_na_cg), stat='identity') +
    xlab('Sample Name') + ylab('Fraction of NAs (%)') +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1))
WGG(p1) + WGG(p2, RightOf())


```


