---
title: "EWAS SVA for cell clustering"
author: "Dan Evans"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  BiocStyle::html_document:
    toc_float: true
    toc_depth: 3
    fig_caption: yes

fontsize: 14 pt

vignette: >
  %\VignetteIndexEntry{EWAS SVA for cell clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

---

# Introduction

EPIC array has type 1 and type 2 probes. Type 1 probes are from the old 27K array that uses 2 bead types per CpG. The type I probes are labeled I-green and I-red. The newer type 2 probes uses one bead type. Most of the EPIC array probes are type 2. 

DNA methylation quantity is expressed as $\beta$ 

$$ \beta = M/(M + U)  $$

Where M = hybridization signal from a methylated version of a cytosine nucleotide and 
U =  hybridization signal from an unmethylated version of a cytosine nucleotide. 

Beta can be interpreted as the proportion of methylation signal for a probe, and values range from 0 to 1. Beta is easy to interpret for humans, but typically has a bimodal distribution that is suboptimal for statistical modeling. Thus, we analyze M-values, which are another way to express methylation values for probes. 

$$ M-value = log_2(M/U) $$

A detection probability represents the probability of a detected signal being background flourescence. If the probability is high, the signal is more likely to be background, and the value should be set to missing.

Standard workflows suggest to remove data points with detection P-value > 0.05.  

```{r, setup}
knitr::opts_chunk$set(cache.lazy = FALSE)
```

## Load libraries
```{r message = FALSE}
library(tidyverse)
library(readxl)
library(knitr)
library(sesame)
library(wheatmap)
library(multtest)
library(limma)
library(RColorBrewer)
library(EnhancedVolcano)
library(kableExtra)
```


# Blood cell type analysis

Does CBC or Horvath predicted cell types associate with PCs?

Also, do Horvath predictions associate with CBC?

```{r cache = TRUE}

eset_Mvals <- read_rds("../data/formatted/eset_Mvals_clean.rds")
f_dat <- fData(eset_Mvals)
p_dat <- pData(eset_Mvals)
e <- exprs(eset_Mvals)
miss_probe <- apply(e, 1, function(x) sum(is.na(x)))
sum(miss_probe > 0)
sum(f_dat$CVprobe >= 100)
sum(f_dat$CVprobe >= 100 & miss_probe == 0)
sum(miss_probe == 0 & f_dat$CVprobe < 100)
eset_sub <- eset_Mvals[miss_probe == 0 & f_dat$CVprobe < 100, ] 
eset_pca <- prcomp(t(exprs(eset_sub)), center = TRUE, scale = TRUE)
summary(eset_pca)$importance[,1:10]
eset_PC <- eset_pca[["x"]]
eset_PC <- as.data.frame(eset_PC)
eset_PC$ID <- row.names(eset_PC)
eset_PC <- eset_PC %>%
	select(ID, PC1, PC2)

pheno <- read_csv("../data/pheno/INFLAME.CSV")
phenoV1 <- read_csv("../data/pheno/V1FEB14.CSV")
phenoV1 <- phenoV1 %>%
	select(ID, GIRACE, GIWHITE, GIAA, GIASIAN, GIHISPA)
with(phenoV1, table(GIRACE, GIWHITE))
with(phenoV1, table(GIRACE, GIAA))
with(phenoV1, table(GIRACE, GIASIAN))
with(phenoV1, table(GIRACE, GIHISPA))
pheno <- pheno %>%
	inner_join(phenoV1, by = "ID")
pheno <- pheno %>%
	inner_join(p_dat, by = c("ID" = "Sample_Name"))
with(pheno, table(GIRACE))
pheno %>%
	filter(GIRACE != 1) %>%
	select(ID, GIRACE)

horvath <- read_csv("../data/horvath/output/MethylData.output.csv")
table(horvath$predictedGender)

eset_PC <- eset_PC %>%
	inner_join(pheno, by = "ID")

summary(eset_PC$LABASO) #basophils 
table(cut(eset_PC$LABASO, breaks = 4))
summary(eset_PC$LAEOSI) #eosinophils 
table(cut(eset_PC$LAEOSI, breaks = 4))
summary(eset_PC$LALYMP) #lymphocytes 
table(cut(eset_PC$LALYMP, breaks = 4))
summary(eset_PC$LAMONO) #monocytes 
table(cut(eset_PC$LAMONO, breaks = 4))
summary(eset_PC$LANEUT) #neutrophils 
table(cut(eset_PC$LANEUT, breaks = 4))

eset_PC %>%
	ggplot(aes(PC1, PC2)) +
	geom_point(aes(color = cut(LABASO, breaks = 4)))

eset_PC %>%
	ggplot(aes(PC1, PC2)) +
	geom_point(aes(color = cut(LAEOSI, breaks = 4)))

eset_PC %>%
	ggplot(aes(PC1, PC2)) +
	geom_point(aes(color = cut(LALYMP, breaks = 4)))

eset_PC %>%
	ggplot(aes(PC1, PC2)) +
	geom_point(aes(color = cut(LAMONO, breaks = 4)))

eset_PC %>%
	ggplot(aes(PC1, PC2)) +
	geom_point(aes(color = cut(LANEUT, breaks = 4)))
```


